---
title: "Final_Project_Report"
author: "Tong Zhang, Zhiqiang Liao"
date: "August 14, 2015"
output: pdf_document
---
Description: 
  The final Project involves analyzing storms data from NCDC (National Climatic Data Center)'s International Best Track Archive for Climate Stewardship (IBTrACS). We mainly works with data from East Pacific and North Atliantic basins.
  
Data Source: 

The link below consists data of storms from East Pacific: ftp://eclipse.ncdc.noaa.gov/pub/ibtracs/v03r06/wmo/csv/basin/Basin.EP.ibtracs_wmo.v03r06.csv
The link below consists data of storms from North Pacific:
Basin.NA.ibtracs_hurdat.v03r06.hdat
ftp://eclipse.ncdc.noaa.gov/pub/ibtracs/v03r06/wmo/csv/basin/Basin.NA.ibtracs_wmo.v03r06.csv


For data processing section, we produced two csv files that processed the raw data 
file for the North Atlantic basin in HURDAT format. The storms.csv based on the data from the HEADER of each storm. It includes id (stroms number since 1851), date, days (number of days in which positions are available), and name. The tracks.csv based on the HEADER and DAILY DATA from each storm. It includes id, date, periond (00h, 06h,12h, 18h), stage (type of stage), lat (latitude), long (longitude), wind (wind speed), and press (pressure).
  
  For Data Analysis section, we focused on analyzing data of the North Atlantic Basin from 1980 to 2010. We used the previous cleaned-processed files storms.csv and tracks.csv as sources.
  
```{r, echo= FALSE, message = FALSE, warning = FALSE}
# =============================================================================
library(stringr)
library(dplyr)
library(maps)
library(ggplot2)
storms<-read.csv('../data/storms.csv', stringsAsFactors = FALSE)
tracks<-read.csv('../data/tracks.csv', stringsAsFactors = FALSE)

# =============================================================================
# number of storms per year
# =============================================================================

# first extract years from date
date = as.Date(storms$date, '%m/%d/%Y')
year = format(date, '%Y')
year = year[which(year >= 1980 & year <= 2010)]

# find the frequency of storms in each year
stormPerYear<-table(year)
stormPerYear
barplot(as.numeric(stormPerYear),names.arg = unique(year),
        main = 'number of storms per year',
        xlab = 'Year', ylab = 'number of storms')

# =============================================================================
# number of storms per year with wind >=35
# =============================================================================
# Extract years of which the storm's speed exceed 35
# tracks$id[tracks$wind>=35]: find all ids of storms that have wind speed exceed 35 mph.
# unique the id so we know how many storms there are
# find the date of thoses storms that matches the unique(id)
# extract the year from date we can get years which storm wind speed exceed 35
date = as.Date(tracks$date, '%m/%d/%Y')
year = format(date, '%Y')
tmpTracks = tracks[which(year >= 1980 & year <= 2010),]
year35Date = as.Date(storms$date[unique(tmpTracks$id[tmpTracks$wind>=35])], '%m/%d/%Y')
year35 = format(year35Date, "%Y")
# using table to find frequency of storms in each year
stormPerYear35<-table(year35)
stormPerYear35
barplot(as.numeric(stormPerYear35), names.arg=unique(year35), 
        main='number of storms per year with wind >=35',
        xlab = 'Year', ylab = 'number of storms')
# =============================================================================
# number of storms per year with wind >=64
# =============================================================================
# years of which the storm's speed exceed 64
#year64<-str_extract(storms$date[unique(tmpTracks$id[tmpTracks$wind>=64])],'[[:digit:]]{4}')
year64Date = as.Date(storms$date[unique(tmpTracks$id[tmpTracks$wind>=64])], '%m/%d/%Y')
year64 = format(year64Date, "%Y")
# using table to find frequency of storms in each year
stormPerYear64<-table(year64)
stormPerYear64
barplot(as.numeric(stormPerYear64), names.arg=unique(year64),
        main='number of storms per year with wind >= 64',
        xlab = 'Year', ylab = 'number of storms')

# =============================================================================
# number of storms per year with wind >=96
# =============================================================================
# first extract years of which the storm's speed exceed 96
year96Date = as.Date(storms$date[unique(tmpTracks$id[tmpTracks$wind>=96])], '%m/%d/%Y')
year96 = format(year96Date, "%Y")
# using table to find frequency of storms in each year
stormPerYear96<-table(year96)
stormPerYear96
barplot(as.numeric(stormPerYear96), names.arg=unique(year96),
        main='number of storms per year with wind >= 96',
        xlab = 'Year', ylab = 'number of storms')
```

\newpage
```{r, echo=FALSE}
date = as.Date(storms$date, '%m/%d/%Y')
monthWithYear = format(date, '%m/%Y')
year = format(date, '%Y')
monthWithYear = monthWithYear[year >= 1980 & year <= 2010]

# find the frequency of storms that happened in each month
month = substring(text = monthWithYear, first = 1, last = 2)
stormPerMonth<-table(month)
# plot
monthName = month.name[sort(as.numeric(unique(month)))]
stormPerMonth
barplot(as.numeric(stormPerMonth), names.arg = monthName,
        main = 'number of storms per month',
        xlab = 'Year', ylab = 'number of storms')

# =============================================================================
# number of storms per month with wind >=35
# =============================================================================
# tracks$id[tracks$wind>=35]: find all ids of storms that have wind speed exceed 35 mph.
# unique the id so we know how many storms there are that wind speed exceed 35
# find the date of thoses storms that matches the unique(id)
# extract the month from date we can get years which storm wind speed exceed 35 
date = as.Date(tracks$date, '%m/%d/%Y')
monthWithYear = format(date, '%m/%Y')
year = format(date, '%Y')
tmpTracks = tracks[which(year >= 1980 & year <= 2010),]

# find the frequency of storms that happened in each month
month35<-storms$date[unique(tmpTracks$id[tmpTracks$wind>=35])]
month35 = as.Date(month35, '%m/%d/%Y')
month35 = format(month35, '%m')
stormPerMonth35<-table(month35)

# plot
monthName = month.name[sort(as.numeric(unique(month35)))]
stormPerMonth35
barplot(as.numeric(stormPerMonth35), names.arg = monthName,
        main = 'number of storms per month with wind >=35',
        xlab = 'Year', ylab = 'number of storms')

# =============================================================================
# number of storms per month with wind >=64
# =============================================================================
# find the frequency of storms that happened in each month
month64<-storms$date[unique(tmpTracks$id[tmpTracks$wind>=64])]
month64 = as.Date(month64, '%m/%d/%Y')
month64 = format(month64, '%m')
stormPerMonth64<-table(month64)

# plot
monthName = month.name[sort(as.numeric(unique(month64)))]
stormPerMonth64
barplot(as.numeric(stormPerMonth64), names.arg = monthName,
        main = 'number of storms per month with wind >=64',
        xlab = 'Year', ylab = 'number of storms')

# =============================================================================
# number of storms per month with wind >=96
# =============================================================================
# find the frequency of storms that happened in each month
month96<-storms$date[unique(tmpTracks$id[tmpTracks$wind>=96])]
month96 = as.Date(month96, '%m/%d/%Y')
month96 = format(month96, '%m')
stormPerMonth96<-table(month96)

# plot
monthName = month.name[sort(as.numeric(unique(month96)))]
stormPerMonth96
barplot(as.numeric(stormPerMonth96), names.arg = monthName,
        main = 'number of storms per month with wind >=96',
        xlab = 'Year', ylab = 'number of storms')
```
```{r, echo=FALSE, message= FALSE, warning =FALSE}
# =============================================================================
# Annual numbers of storms >= 35/64/96
# =============================================================================
## Average 
mean35<-mean(stormPerYear35)
mean64<-mean(stormPerYear64)
mean96<-mean(stormPerYear96)
## std
std35<-sd(stormPerYear35)
std64<-sd(stormPerYear64)
std96<-sd(stormPerYear96)
## 25 percentile
Q35<-summary(as.numeric(stormPerYear35))[2]
Q64<-summary(as.numeric(stormPerYear64))[2]
Q96<-summary(as.numeric(stormPerYear96))[2]
## median
median35<-summary(as.numeric(stormPerYear35))[3]
median64<-summary(as.numeric(stormPerYear64))[3]
median96<-summary(as.numeric(stormPerYear96))[3]
## 75 percentile
q35<-summary(as.numeric(stormPerYear35))[5]
q64<-summary(as.numeric(stormPerYear64))[5]
q96<-summary(as.numeric(stormPerYear96))[5]

## table
annual<-matrix(c(mean35,std35,Q35, median35,q35,
                 mean64,std64,Q64, median64,q64,
                 mean96,std96,Q96, median96,q96), ncol=5, byrow=TRUE)
colnames(annual)<-c('Avg','Std.Dev','25th','50th','75th')
rownames(annual)<-c('35 knots','64 knots', '96 knots')
annual = signif(annual, 4)
annual<-as.table(annual)
annual
```

\newpage
```{r, echo=FALSE, message= FALSE, warning = FALSE}
# =============================================================================
# Regression Line
# =============================================================================
# mean pressure and mean wind speed for each storm
vec<-c(1:1777)
pressMean1<-sapply(vec, function(x) 
  mean(tmpTracks$press[tmpTracks$id==x]))
# remove zeros
pressMean<-pressMean1[pressMean1!=0]
# mean wind speed
windMean1<-sapply(vec,function(x) 
  mean(tmpTracks$wind[tmpTracks$id==x]))
#remove zeros
windMean<- windMean1[pressMean1!=0]
reg = lm(pressMean ~ windMean)
plot(windMean, pressMean,xlab = 'mean wind speed', ylab = 'mean pressure')
title('Mean Wind VS Mean Wind Speed for Each Storm ')
abline(reg)

# =============================================================================
# Regression analysis 2
# =============================================================================
# median pressure and median wind speed for each storm
vec<-c(1:1777)
pressMedian1<-sapply(vec,function(x) 
  median(tmpTracks$press[tmpTracks$id==x]))
# mean wind speed
windMedian1<-sapply(vec,function(x) 
  median(tmpTracks$wind[tmpTracks$id==x]))
# remove zeros
pressMedian<-pressMedian1[pressMedian1!=0]
windMedian<-windMedian1[pressMedian1!=0]

reg1 = lm(pressMedian ~ windMedian)
plot(windMedian, pressMedian, xlab = 'Median wind speed', ylab = 'Median pressure')
title('Median Wind Speed VS Median Pressure For Each Storm ')
abline(reg1)

```

\newpage
For the visualization part of the project involves using data to visualize the trajectory of the storms in both the East Pacific and the North atliantic basins from 1980 to 2010. 

```{r, echo= FALSE,message=FALSE, warning=FALSE}

# =============================================================================
# loading data
# =============================================================================
EastP<-read.csv('../rawdata/EastP.csv',skip = 1, header = TRUE)
NorthA<-read.csv('../rawdata/NorthA.csv', skip=1, header=TRUE)


# =============================================================================
# Preprocessing Data
# =============================================================================

cleanEP<-EastP[(as.character(EastP$Season)>='1980' & 
                as.character(EastP$Season)<= '2010'),]
cleanNA<-NorthA[(as.character(NorthA$Season)>='1980' & 
                as.character(NorthA$Season)<= '2010'),]


## change the format in each column
cleanEP$Latitude<-as.numeric(str_trim(as.character(cleanEP$Latitude), side='left'))
cleanEP$Longitude<-as.numeric(str_trim(as.character(cleanEP$Longitude), side='left'))
cleanEP$Wind.WMO.<-as.numeric(str_trim(as.character(cleanEP$Wind.WMO.), side='left'))


## change the format in each column in North America Basin
cleanNA$Latitude<-as.numeric(str_trim(as.character(cleanNA$Latitude), side='left'))
cleanNA$Longitude<-as.numeric(str_trim(as.character(cleanNA$Longitude), side='left'))
cleanNA$Wind.WMO.<-as.numeric(str_trim(as.character(cleanNA$Wind.WMO.), side='left'))

## Combine the two data frame
# clean<-rbind(cleanEP, cleanNA)
clean<- full_join(cleanEP, cleanNA)

## change Serial_Num column to factor
clean$Serial_Num<-as.factor(clean$Serial_Num)

## Extract Month
clean$Month<-str_replace_all(str_extract(as.character(clean$ISO_time),
            '-[0-9]{2}-'),'-','')
clean$Month<-factor(clean$Month, labels=month.name)

# gather world map data from maps package
mapdata<-map_data('world')
```

Visulization 1: Trajectory of storms from 1980 to 2010

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=12, fig.height=12}
globalmap<- ggplot(clean, aes(x=Longitude, y=Latitude, group=Serial_Num)) +
  geom_polygon(data=mapdata, aes(x=long, y=lat, group=group),
               fill='#000050', colour='#000050', size=0.5) +
  geom_path(data=clean, aes(group=Serial_Num, colour=Wind.WMO.), alpha=0.5, size=0.5) +
  ggtitle('Trajectory of storms from 1980 to 2010')+
  scale_color_continuous(name='Wind (knot)',low = '#E68A00',high = '#4C2E00',
                         space = 'Lab',guide = 'colourbar')+
  xlim(-150, -25) + ylim(5, 60)+
  theme(panel.background = element_rect(fill='#00006BDD',colour='#00006BDD'),
        panel.grid = element_blank(),
        panel.border = element_blank())

library(png)
library(grid)
grid.raster(readPNG("../images/globalMap.png"))

```

\newpage
Visulization 2: Trajectory of storms per month from 1980 to 2010
(one facet per month)

```{r, echo=FALSE, message= FALSE, warning= FALSE, fig.width=12, fig.height=12}
stormbymonth <- ggplot(clean, aes(x=Longitude, y=Latitude, group=Serial_Num)) +
  geom_polygon(data=map_data("world"), aes(x=long, y=lat, group=group),
               fill='#000050', colour='#000050', size=0.5) +
  geom_path(data=clean, aes(group=Serial_Num, colour=Wind.WMO.), 
            alpha=0.2,size=0.5) +
  ggtitle('Trajectory of storms per month from 1980 to 2010')+
  scale_color_continuous(name='Wind (knot)',low = '#E68A00',high = '#4C2E00',
                         space = 'Lab',guide = 'colourbar')+
  xlim(-150, -25) + ylim(5, 60)+facet_wrap(~ Month)+
  theme(panel.background = element_rect(fill='#00006BDD',colour='#00006BDD'),
        panel.grid = element_blank(),
        panel.border = element_blank())

grid.raster(readPNG("../images/stormByMonth.png"))
```

\newpage
Visulization 3: trajectory of storms in decade 1980s 
(one facet per year)

```{r, echo = FALSE, message = FALSE, warning = FALSE, fig.width=12, fig.height=12}
# extract storms from 1980 to 1989
traj1980<-clean[as.character(clean$Season)>='1980' & 
                as.character(clean$Season)<='1989',]

# As. factor Season column in the data frame
traj1980$Season<-as.factor(traj1980$Season)

# plot map
map1980 <- ggplot(traj1980, aes(x=Longitude, y=Latitude, group=Serial_Num)) +
  geom_polygon(data=map_data("world"), aes(x=long, y=lat, group=group),
               fill='#000050', colour='#000050', size=0.5) +
  geom_path(data=traj1980, aes(group=Serial_Num, colour=Wind.WMO.),alpha=0.2, size=0.5) +
  ggtitle('Trajectory of storms per month from 1980 to 1989')+
  scale_color_continuous(name='Wind (knot)',low = '#E68A00',high = '#4C2E00',
                         space = 'Lab',guide = 'colourbar')+
  xlim(-150, -25) + ylim(5, 60)+facet_wrap(~ Season)+
  theme(panel.background = element_rect(fill='#00006BDD',colour='#00006BDD'),
        panel.grid = element_blank(),
        panel.border = element_blank())

grid.raster(readPNG("../images/stormsIn1980s.png"))
```

\newpage
Visulization 4: trajectory of storms in decade 1990s 
(one facet per year)

```{r, echo = FALSE, message = FALSE, warning = FALSE, fig.width=12, fig.height=12}
# Extract storms from 1990 to 1999
traj1990<-clean[as.character(clean$Season)>='1990' & 
                as.character(clean$Season)<='1999',]

# As. factor Season column in the storm data frame
traj1990$Season<-as.factor(traj1990$Season)

# plot map
map1990 <- ggplot(traj1990, aes(x=Longitude, y=Latitude, group=Serial_Num)) +
  geom_polygon(data=map_data("world"), aes(x=long, y=lat, group=group),
               fill='#000050', colour='#000050', size=0.5) +
  geom_path(data=traj1990, aes(group=Serial_Num, colour=Wind.WMO.), 
            alpha=0.2, size=0.5)+
  ggtitle('Trajectory of storms per month from 1990 to 1999')+
  scale_color_continuous(name='Wind (knot)',low = '#E68A00',high = '#4C2E00',
                         space = 'Lab',guide = 'colourbar')+
  xlim(-150, -25) + ylim(5, 60)+facet_wrap(~ Season)+
  theme(panel.background = element_rect(fill='#00006BDD',colour='#00006BDD'),
        panel.grid = element_blank(),
        panel.border = element_blank())

grid.raster(readPNG("../images/stormsIn1990s.png"))

```

\newpage
Visulization 5: trajectory of storms in decade 2000s 
(one facet per year)

```{r, echo = FALSE, message = FALSE, warning = FALSE, fig.width=12, fig.height=12}

# Extract storms from 2000 to 2009
traj2000<-clean[as.character(clean$Season)>='2000' & 
                as.character(clean$Season)<='2009',]

# As. factor Season column in the storm data frame
traj2000$Season<-as.factor(traj2000$Season)

# plot map
map2000 <- ggplot(traj2000, aes(x=Longitude, y=Latitude, group=Serial_Num)) +
  geom_polygon(data=map_data("world"), aes(x=long, y=lat, group=group),
               fill='#000050', colour='#000050', size=0.5) +
  geom_path(data=traj2000, aes(group=Serial_Num, colour=Wind.WMO.),
            alpha=0.2, size=0.5) +
  ggtitle('Trajectory of storms per month from 2000 to 2009')+
  scale_color_continuous(name='Wind (knot)',low = '#E68A00',high = '#4C2E00',
                         space = 'Lab',guide = 'colourbar')+
  xlim(-150, -25) + ylim(5, 60)+facet_wrap(~ Season)+
  theme(panel.background = element_rect(fill='#00006BDD',colour='#00006BDD'),
        panel.grid = element_blank(),
        panel.border = element_blank())

grid.raster(readPNG("../images/stormsIn2000s.png"))
```

Extra Credit:

Github: https://github.com/isaaclzq/stat133_project.git

LinkedIn: https://www.linkedin.com/pulse/storm-project-tong-zhang?trk=prof-post
